---
title: "Oregon Bees and Wildfires"
author: "Ali Carmichael"
date: "2024-12-08"
output: pdf_document
---
Introduction:

  In Oregon, recent data shows a clear increase in wildfire frequency and intensity, largely due to extreme heat, drought frequency, and less cold weather caused by climate change. Increased temperature and lack of moisture in fire fuel like trees, shrubs, grasses, and forest debris can create several factors required to fuel large-scale fires. According to Oregon Climate Assessments, Oregon saw more days per year near 90 degrees between 2011 and 2020 than between 1951 and 2010 (Rojas, 2023). Since 2015, Oregon has experienced significant, record-breaking wildfire seasons that have led to massive destruction and ecosystem disruptions. A study cited by the National Oceanic and Atmospheric Administration (NOAA, 2023) found that climate change has doubled the area burned in western U.S. forests between 1984 and 2015. Furthermore, research shows that an average annual increase in temperature of 1 degree Celsius could increase the median burned area per year by as much as 600% in some forests (NOAA, 2023). Ongoing warming could continue amplifying fire seasons, posing long-term threats to Oregon’s forest and wildlife. 
  According to the USDA, pollinators such as bees, butterflies, and birds pollinate around three-fourths of the world’s flowering plants, making pollinators essential for biodiversity (USDA Pollinators). The increase in wildfires, especially in regions like Oregon, poses a significant threat to pollinators. Wildfires are responsible for habit loss that has contributed to the decline of many pollinator species (USDA Pollinators, 2024). Fire can disrupt the balance of ecosystems and reduce plant diversity that pollinator species, especially bees, rely on to survive. In some cases, natural wildfires can benefit pollinators by promoting the growth of smaller, low-growing, flowering plants that support diverse bee populations. However, with the increase of recent wildfire activity, any potential benefit from these fires may be outweighed by the harm they cause to pollinator species, especially bees. 
	Similar to bees, when wildfires burn at intervals that allow for enough recovery time ecosystems can benefit from the effects. Fires destroy leaf litter, dead wood, and other ground cover, triggering forests’ rebirth (USDA Forest Service). Many plant species have adapted to protect themselves against fire and some species even benefit from periodic fires, like sun-loving flowering plants or shrubs. However, the increase in high-intensity fires threatens Oregon’s forest ecosystems. Without adequate recovery time between fires we may begin to see a decrease in species that rely on old-growth forests, and large older trees that lived through low-intensity burns may also begin to disappear (Adlam et al., 2022). In addition, water that flows over burned areas collects nutrients and sediments, washing them downstream where native plant species will not be able to access them to grow (USDA Forest Service). Invasive plants also share many of the characteristics needed to grow after a fire. They will often out compete native plants, displacing them and causing soil erosion as well as a reduction in habitat for wildlife (USDA Forest Service).
  Our study looks at the recovery of bee and native plant populations in the area affected by the Santiam fire in 2020. We selected the Santiam fire because it was composed of three separate fires in the same area during the same time interval. The Santiam fire consists of three fires, the Lionshead fire, Riverside fire, and Beachie Creek fire. To investigate recovery patterns, we will analyze data from the Oregon Bee Atlas, which includes details on bee species, their locations, and the plants they associate with (Oregon Bee Project, 2023). This information is essential for identifying which bee species recover first and the specific plants they are associated with in post-fire zones. Fire zone data from the Monitoring Trends Burn Severity (MTBS) will be used to identify burned areas. By combining these data sets, we want to assess changes in bee and native plant species diversity before and after wildfires by calculating the species richness. This study will help bring attention to the ecological endangerment of bee populations and provide bee conservationists with insights to better assist our pollinating friends in becoming more resilient to wildfires.
  Pollinators, especially bees, are vital to ecosystem health and especially agricultural productivity. Bees play an essential role in the pollination of many plants, including crops. As discussed earlier, bees are in charge of pollinating a little over a third of US crops (USDA, 2024). The heavy reliance on pollinators means we are obligated to protect them. According to the Environmental Defense Fund, As climate change intensifies, extreme heat and drought conditions are leading to more frequent and severe wildfires. These fires cause widespread destruction to habitats, affecting both flora and fauna and threatening the survival of pollinator populations (Environmental Defense Fund, 2024). Given the growing ecological threats caused by climate change, there is an urgent need to understand how bee populations recover in post-fire environments and identify factors that support resilience. Visualizing recovery patterns across different habitats using spatial data can provide helpful insights into which ecosystems and native plants foster quicker bee recovery. 
	The goal of this project is to answer the following question about how wildfires affect bee populations and the flora they interact with: How do bee and Native plant species differ in areas before and after wildfires? We expect the composition of bee and native plant species in a post-fire setting to differ significantly from that of a pre-fire setting. We also predict that the bee species in post-fire areas will associate with early-successional plant species that can colonies and then transmit to more diverse flora as recovery accelerates.

Packages used:
```{r, Load in packages used}
library(terra)
library(ggplot2)
library(dplyr)
library(sf)
library(maps)
library(ggspatial)
library(plotly)
library(knitr)
library(stringr)
library(vegan)
library(igraph)
library(networkD3)
```


Data:
The data sets used for this project were obtained from the Oregon Bee Atlas database. These data sets are comprised of iNaturalist observations of bee species and associated plant species. We also used three Shapefiles of the Lions Head, Beachie Creak, and Riverside Fire to create burn borders for the 2020 Santiam Fire. These data sets were downloaded from the Monitoring Trends Burn Severity (MTBS) website and plotted onto an Oregon map.

```{r, Load in datasets and cleaning procedures}

# Load in Oregon Bee Atlas data
oba <- read.csv("project/data/OBA_2018-2023.csv") 

#load in fire border data
LionsHead_sf <- read_sf("project/data/LionsHead/or4472312167920200817_20200810_20210810_burn_bndy.shp")
BeachieCreek_sf <- read_sf("project/data/BeachieCreek/or4482112218820200816_20200810_20210810_burn_bndy.shp")
Riverside_sf <- read_sf("project/data/Riverside/or4504912206220200908_20190720_20210725_burn_bndy.shp")

# Load Oregon county map data
oregon_map <- map_data("county", "oregon") %>% 
  select(long, lat, group, id = subregion)  

#load in preview of datasets
head(oba)
head(LionsHead_sf)
head(BeachieCreek_sf)
head(Riverside_sf)
```

Cleaning Procedures of OBA data:
Although the shape files for the fire borders required no cleaning, the Oregon Bee Atlas data had to be cleaned to create clear and consistent bee/plant species names. We also needed to remove some rows that had missing observations that didn't include both bee and plant observation. Lastly we created a new column "Genus Species" where we merged columns "genus" and "species" together.
```{r,Cleaning Procedures of OBA data}
#We have a some rows with blank values or spaces that make the data difficult to use, we will remove these to make analysis easier.

#removed blanks
final_oba_df <- final_oba_df[final_oba_df$Associated.plant...genus..species != "",]
final_oba_df <- final_oba_df[final_oba_df$GenusSpecies != " ",]
is.na(final_oba_df)
final_oba_df


#We also need to fix all the weird species names for the plants. There are a lot of weird names that make analysis challenging.

## Remove the special character
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, "\x92", "")

# Removing blanks
final_oba_df <- final_oba_df[final_oba_df$Associated.plant...genus..species != "",]

# Removing "net"
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, "net", "")

# Removing yarrow
final_oba_df$Associated.plant...genus..species[final_oba_df$Associated.plant...genus..species == "Yarrow"] <- "Achillea millefollium"

# Removing non plant
final_oba_df <- final_oba_df[final_oba_df$Associated.plant...genus..species != "Weedy yellow comp.",]

# Removing common names in ()
final_oba_df$Associated.plant...genus..species <- gsub("\\s*\\(.*?\\)", "", final_oba_df$Associated.plant...genus..species)

# Removing weird words
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, ",.*$", "")

# Removing space
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, "\\s+$", "")

# Removing "sp."
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, " sp$", " sp.")

# Removing "or"
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, " or .*", "")

# Removing lists
final_oba_df <- final_oba_df[!grepl(",", final_oba_df$Associated.plant...genus..species),]

head(final_oba_df)

#Here we are creating two columns to sort the plant and bee species by genus and species. When we make our networks it will make it easier to have just the genus and species sorted. We will start with identifying the species resolution then we will drop the values that are missing a genus or species.

# Initialize the empty columns to NA values
final_oba_df$Plant.resolution <- NA
final_oba_df$Bee.resolution <- NA

# Loop through the data frame and then check if there is only one word in the
# plant column. If there's one word it will be the Genus, otherwise it's species.
for (i in 1:nrow(final_oba_df)) {
  if (str_count(final_oba_df$Associated.plant...genus..species[i], "\\S+") == 1) {
    final_oba_df$Plant.resolution[i] <- "Genus"
  }
  else {
    final_oba_df$Plant.resolution[i] <- "Species"
  }
}

# Similar to above we will loop through our data frame and determine if the Bee resolution is
# Genus or Species
for (i in 1:nrow(final_oba_df)) {
  if (str_count(final_oba_df$GenusSpecies[i], "\\S+") == 1) {
    final_oba_df$Bee.resolution[i] <- "Genus"
  }
  else {
    final_oba_df$Bee.resolution[i] <- "Species"
  }
}
```

Data Wrangling and Summarizing:
```{r}

```

Hypothesis Tests:
(Paragraph discussing the Tests used)
```{r, Load i all tests used}

```

Visualization:
```{r, load in all final visualizations}

```


Discussion/Conclusion:

"No need to load in code for this section"


Load in the Oregon Bee Atlas data using the read.csv() function.
```{r}
# Load in Oregon Bee Atlas data

oba <- read.csv("project/data/OBA_2018-2023.csv") 
head(oba)
```
Checking how many NA Values we have in our Latitude and Longitude column. We won't be able 
to use rows with NA's so we need to drop them.
```{r}
# Initial check
sum(is.na(oba$Dec..Lat.))
sum(is.na(oba$Dec..Long.))

# Filter out the NA values
oba <- oba %>% 
  filter(!is.na(Dec..Lat.) & !is.na(Dec..Long.))

# Final Check
sum(is.na(oba$Dec..Lat.))
sum(is.na(oba$Dec..Long.))
```
Loading in the a map of Oregon for plotting our observations.
```{r, Oregon Map}
# Load Oregon county map data
oregon_map <- map_data("county", "oregon") %>% 
  select(long, lat, group, id = subregion)  # Correct column name is `region`
```

Plotted the bee observations so we can see what areas aren't reached by OBA. It also shows us the distribution of our
observations and if we have any that seem like they might be incorrect. This is just an intial visualization to explore
the observation points.
```{r graph bee data}
oba_observations_sf <- st_as_sf(oba, coords = c("Dec..Long.", "Dec..Lat."), crs = 4326)

#st_crs(oba_observations_sf)
#st_crs(oregon_map_sf) == st_crs(oba_observations_sf)

#plot bee observations on to oregon map with fires
ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey", color = "white") +
  geom_sf(data = oba_observations_sf, aes(color ="blue"), size = .1, alpha = 0.9) +
  labs(title = "Bee Species Observations in Oregon",
       x = "Longitude", 
       y = "Latitude") +
  theme_minimal() +
  coord_sf()

```

Next, we loaded in our fire data sets. These are three fires that occurred in the same area in 2020. Two of the fire
lines merged and the other fire was really close to the borders of the merged fire. These three fires are often
recorded as one large fire called the Santiam fire. We are using these because they were 3 very large fires in 2020 
that have a lot of bee observations around them. 
```{r}
# Load in each fire shape file
LionsHead_sf <- read_sf("project/data/LionsHead/or4472312167920200817_20200810_20210810_burn_bndy.shp")

BeachieCreek_sf <- read_sf("project/data/BeachieCreek/or4482112218820200816_20200810_20210810_burn_bndy.shp")

Riverside_sf <- read_sf("project/data/Riverside/or4504912206220200908_20190720_20210725_burn_bndy.shp")

# Set a common CRS 
target_crs <- st_crs(4326)

# Transform each fire to the target CRS
LionsHead_sf <- st_transform(LionsHead_sf, target_crs)
BeachieCreek_sf <- st_transform(BeachieCreek_sf, target_crs)
Riverside_sf <- st_transform(Riverside_sf, target_crs)

#check if they match
print(st_crs(LionsHead_sf) == st_crs(BeachieCreek_sf))
print(st_crs(BeachieCreek_sf) == st_crs(Riverside_sf))

# Plotting the fires on the oregon map loaded in earlier
ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey", color = "white") +
  geom_sf(data = LionsHead_sf, color = "Red", alpha = 0.9) +
  geom_sf(data = BeachieCreek_sf, color = "Red", alpha = 0.9) +
  geom_sf(data = Riverside_sf, color = "Red", alpha = 0.9) +
  labs(
    title = "Distribution of Oregon Fires in 2020",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  coord_sf()

#Now lets zoom in

# merge all fires into one
fire_combined_bbox <- st_bbox(rbind(LionsHead_sf, BeachieCreek_sf, Riverside_sf))

# Adjust the margins 
fire_combined_bbox <- fire_combined_bbox + c(-0.1, -0.1, 0.1, 0.1)  #(xmin, ymin, xmax, ymax)

# Final plotting of the zoomed in section of Oregon
ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey", color = "white") +
  geom_sf(data = LionsHead_sf, aes(color = "Lions Head"), alpha = 0.9) +
  geom_sf(data = BeachieCreek_sf, aes(color = "Beachie Creek"), alpha = 0.9) +
  geom_sf(data = Riverside_sf, aes(color = "Riverside"), alpha = 0.9) +
  scale_color_manual(
    name = "Fires",
    values = c("Lions Head" = "red", "Beachie Creek" = "purple", "Riverside" = "yellow")
  ) +
  labs(
    title = "Zoomed-In View of Oregon Fires in 2020",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  coord_sf(
    xlim = c(fire_combined_bbox["xmin"], fire_combined_bbox["xmax"]),
    ylim = c(fire_combined_bbox["ymin"], fire_combined_bbox["ymax"])
  )
```
Now we want to crop our bee observations to just the ones that are in the borders of the three fires.
We are going to paste our fires into one polygon as it is often recorded as one fire. We used the function 
st_intersection to crop our data set to the fire borders. 
```{r}
#Merge all fires into one object
all_fire_boarder <- rbind(LionsHead_sf, BeachieCreek_sf, Riverside_sf)

#Create a subset of obersvations inside the border
only_fire_bees <- st_intersection(oba_observations_sf, all_fire_boarder)
head(only_fire_bees)

only_fire_bees_df <- as.data.frame(only_fire_bees)

#merges genus and species and makes a new column
only_fire_bees_df$GenusSpecies <- paste(only_fire_bees_df$Genus, only_fire_bees_df$Species)
#create a subset
final_oba_df <- only_fire_bees_df %>% select("Year.1", "GenusSpecies","Associated.plant...genus..species", "BurnBndLat", "BurnBndLon",)

head(final_oba_df)
```

Here we plot our fires and our cropped bee observations from above on the map of Oregon
```{r, subset data}

 ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "tan", color = "black") +
  geom_sf(data = LionsHead_sf, aes(color = "Lions Head"), alpha = 0.9) +
  geom_sf(data = BeachieCreek_sf, aes(color = "Beachie Creek"), alpha = 0.9) +
  geom_sf(data = Riverside_sf, aes(color = "Riverside"), alpha = 0.9) +
  geom_sf(data = only_fire_bees, aes(color = "Bee Observation"), size = 1, alpha = 0.9) +
  scale_color_manual(
    name = "Fire and Observations",
    values = c(
      "Lions Head" = "red",
      "Beachie Creek" = "purple",
      "Riverside" = "Dark Green",
      "Bee Observation" = "blue"
    ),
    breaks = c(
      "Lions Head",
      "Beachie Creek",
      "Riverside",
      "Bee Observation"
    )
  ) + 
  labs(
    title = "Oregon Santiam Fire in 2020",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.background = element_rect(fill = "white")
  ) +
  coord_sf(
    xlim = c(fire_combined_bbox["xmin"], fire_combined_bbox["xmax"]),
    ylim = c(fire_combined_bbox["ymin"], fire_combined_bbox["ymax"])
  )

#ggsave("oregon_fire_map.png", dpi = 300, width = 12, height = 8)  
```

We have a some rows with blank values or spaces that make the data difficult to use,
we will remove these to make analysis easier.
```{r}
#removed blanks
final_oba_df <- final_oba_df[final_oba_df$Associated.plant...genus..species != "",]
final_oba_df <- final_oba_df[final_oba_df$GenusSpecies != " ",]
is.na(final_oba_df)
final_oba_df
```
We also need to fix all the weird species names for the plants. There are a lot of
weird names that make analysis challenging.
```{r}
## Remove the special character
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, "\x92", "")

# Removing blanks
final_oba_df <- final_oba_df[final_oba_df$Associated.plant...genus..species != "",]

# Removing "net"
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, "net", "")

# Removing yarrow
final_oba_df$Associated.plant...genus..species[final_oba_df$Associated.plant...genus..species == "Yarrow"] <- "Achillea millefollium"

# Removing non plant
final_oba_df <- final_oba_df[final_oba_df$Associated.plant...genus..species != "Weedy yellow comp.",]

# Removing common names in ()
final_oba_df$Associated.plant...genus..species <- gsub("\\s*\\(.*?\\)", "", final_oba_df$Associated.plant...genus..species)

# Removing weird words
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, ",.*$", "")

# Removing space
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, "\\s+$", "")

# Removing "sp."
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, " sp$", " sp.")

# Removing "or"
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, " or .*", "")

# Removing lists
final_oba_df <- final_oba_df[!grepl(",", final_oba_df$Associated.plant...genus..species),]

head(final_oba_df)
```

Here we are creating two columns to sort the plant and bee species by genus and species.
When we make our networks it will make it easier to have just the genus and species sorted. We will start
with identifying the species resolution then we will drop the values that are missing a genus or species.
```{r}
# Initialize the empty columns to NA values
final_oba_df$Plant.resolution <- NA
final_oba_df$Bee.resolution <- NA

# Loop through the data frame and then check if there is only one word in the
# plant column. If there's one word it will be the Genus, otherwise it's species.
for (i in 1:nrow(final_oba_df)) {
  if (str_count(final_oba_df$Associated.plant...genus..species[i], "\\S+") == 1) {
    final_oba_df$Plant.resolution[i] <- "Genus"
  }
  else {
    final_oba_df$Plant.resolution[i] <- "Species"
  }
}

# Similar to above we will loop through our data frame and determine if the Bee resolution is
# Genus or Species
for (i in 1:nrow(final_oba_df)) {
  if (str_count(final_oba_df$GenusSpecies[i], "\\S+") == 1) {
    final_oba_df$Bee.resolution[i] <- "Genus"
  }
  else {
    final_oba_df$Bee.resolution[i] <- "Species"
  }
}

# Check our resolutions
table(final_oba_df$Bee.resolution)
table(final_oba_df$Plant.resolution)

# Now we want to filter our plant species to the rows that contain 
# just the genus
final_oba_df$Plant.genus <- ifelse(
  final_oba_df$Plant.resolution %in% c("Genus", "Species"),
  word(final_oba_df$Associated.plant...genus..species, 1), NA)

# Check the plant genus
table(final_oba_df$Plant.genus)

# Now we will filter our bee species to both Genus and species
final_oba_df$Bee.genus <- ifelse(
  final_oba_df$Bee.resolution %in% c("Genus", "Species") & str_count(final_oba_df$GenusSpecies, "\\S+") > 1, word(final_oba_df$GenusSpecies, 1, 2), NA)

# Check our bee genus 
table(final_oba_df$Bee.genus)
```

Drop the NA values to get our final clean data set with the correct bee and plant resolutions.
```{r}
final_oba_df <- final_oba_df %>% 
  filter(!is.na(Plant.genus) & !is.na(Bee.genus))

final_oba_df
```

Add the site status column to our final clean data frame. We want the site status to be "Before" if
the year of the observation was 2020 and before and "After" if the observation was after 2020.
```{r}
# Initialize an empty site status column
final_oba_df$Site.status <- NA

# Loop through the oba data frame
for (i in 1:nrow(final_oba_df)) {
  if (final_oba_df$Year.1[i] <= 2020) {
    final_oba_df$Site.status[i] <- "Before"
  }
  else {
    final_oba_df$Site.status[i] <- "After"
  }
}

# Check our new data frame
head(final_oba_df)
```

Creating a species interaction network for before the fires.
```{r}
# Filter data for 'before' fire interactions
before_fire_interactions <- final_oba_df[final_oba_df$Site.status == "Before", ]

# Create the nodes
nodes_before <- data.frame(
  name = unique(c(before_fire_interactions$Bee.genus, before_fire_interactions$Plant.genus))
)

# Create the links
links_before <- data.frame(
  source = match(before_fire_interactions$Bee.genus, nodes_before$name) - 1,
  target = match(before_fire_interactions$Plant.genus, nodes_before$name) - 1,
  value = 1
)

# Aggregate interactions for unique source-target pairs
links_before <- aggregate(value ~ source + target, data = links_before, sum)

# Create the Sankey network
library(networkD3)

# Define a color scale for nodes
color_scale <- 'd3.scaleOrdinal()
                 .domain(["Bee", "Plant"])
                 .range(["#1f78b4", "#33a02c"])'

sankey_before <- sankeyNetwork(
  Links = links_before,
  Nodes = nodes_before,
  Source = "source",
  Target = "target",
  Value = "value",
  NodeID = "name",
  units = "Interactions",
  fontSize = 14,
  nodeWidth = 40,
  colourScale = color_scale
)

# Here is the network
sankey_before
```

Now we want to calculate some statistics to simualte. We want to calculate the species richness before
and after the fires and then take the difference in richness for both plants and bees.
```{r}
# Richness for bees
Bee_richness_before <- length(unique(final_oba_df$Bee.genus[final_oba_df$Site.status == "Before"]))
Bee_richness_after <- length(unique(final_oba_df$Bee.genus[final_oba_df$Site.status == "After"]))
Bee_richness_change <- Bee_richness_after - Bee_richness_before

# Richness for plants
Plant_richness_before <- length(unique(final_oba_df$Plant.genus[final_oba_df$Site.status == "Before"]))
Plant_richness_after <- length(unique(final_oba_df$Plant.genus[final_oba_df$Site.status == "After"]))
Plant_richness_change <- Plant_richness_after - Plant_richness_before
```

Now we will conduct our simulation using bootstrapping. Our sample size for our "after" group is really small 
so we will use resampling and then a density histogram to simulate what our hypothesis could look
like if there was more data available.
```{r}
# Observed difference in richness
bee_observed_diff <- Bee_richness_after - Bee_richness_before

# Resampling procedure because our "after" group is extremmely small
set.seed(42)  # Set a seed for reproducibility
num_iterations <- 10000  # Increase iterations for a larger distribution

# Perform simulation with shuffled Site.status
bee_null_distribution <- replicate(num_iterations, {
  # Shuffle Site.status
  shuffled_status <- sample(final_oba_df$Site.status)
  
  # Assign the shuffled status to a new column
  final_oba_df$Shuffled.status <- shuffled_status
  
  # Sample from the shuffled "Before" and "After" groups
  bee_sampled_before <- final_oba_df$Bee.genus[final_oba_df$Shuffled.status == "Before"]
  bee_sampled_after <- final_oba_df$Bee.genus[final_oba_df$Shuffled.status == "After"]
  
  # Calculate richness
  bee_richness_before <- length(unique(bee_sampled_before))
  bee_richness_after <- length(unique(bee_sampled_after))
  
  # Return the difference in richness
  bee_richness_after - bee_richness_before
})

# Compare observed difference with null distribution
p_value <- mean(abs(bee_null_distribution) >= abs(bee_observed_diff))
print(paste("p-value: ", p_value))

# Plot the null distribution
bee_null_df <- data.frame(bee_null_distribution)

ggplot(bee_null_df, aes(x = bee_null_distribution)) +
  geom_density(fill = "lightblue", alpha = 0.7) +
  geom_vline(aes(xintercept = bee_observed_diff), color = "red", linetype = "dashed", size = 1) +
  labs(title = "Null Distribution of Bee Species Richness",
       x = "Difference in Richness",
       y = "Density") +
  theme_minimal() +
   theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

#ggsave("simulated_bee_species_richness.png", 
       #plot = bee_plot,
       #dpi = 300,
       #units = "in",
      #bg = "white")
```

Here we conduct our simulation for the plants.
```{r}
# Observed difference in richness
plant_observed_diff <- Plant_richness_after - Plant_richness_before

# Resampling procedure because our "after" group is extremmely small
set.seed(42)  # Set a seed for reproducibility
num_iterations <- 10000  # Increase iterations for a larger distribution

# Perform simulation with shuffled Site.status
plant_null_distribution <- replicate(num_iterations, {
  # Shuffle Site.status
  shuffled_status <- sample(final_oba_df$Site.status)
  
  # Assign the shuffled status to a new column
  final_oba_df$Shuffled.status <- shuffled_status
  
  # Sample from the shuffled "Before" and "After" groups
  plant_sampled_before <- final_oba_df$Plant.genus[final_oba_df$Shuffled.status == "Before"]
  plant_sampled_after <- final_oba_df$Plant.genus[final_oba_df$Shuffled.status == "After"]
  
  # Calculate richness
  plant_richness_before <- length(unique(plant_sampled_before))
  plant_richness_after <- length(unique(plant_sampled_after))
  
  # Return the difference in richness
  plant_richness_after - plant_richness_before
})

# Compare observed difference with null distribution
p_value <- mean(abs(plant_null_distribution) >= abs(plant_observed_diff))
print(paste("p-value: ", p_value))

# Plot the null distribution
plant_null_df <- data.frame(plant_null_distribution)
ggplot(plant_null_df, aes(x = plant_null_distribution)) +
  geom_density(fill = "lightblue", alpha = 0.7) +
  geom_vline(aes(xintercept = plant_observed_diff), color = "red", linetype = "dashed", size = 1) +
  labs(title = "Null Distribution of Plant Species Richness",
       x = "Difference in Richness",
       y = "Density") +
  theme_minimal() +
   theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

#ggsave("simulated_plant_species_richness.png", 
       #dpi = 300,
       #units = "in",
      #bg = "white")
```

Species interaction network for after the fires.
```{r}
# Filter data for 'before' fire interactions
after_fire_interactions <- final_oba_df[final_oba_df$Site.status == "After", ]

# Create the nodes
nodes_after <- data.frame(
  name = unique(c(after_fire_interactions$Bee.genus, after_fire_interactions$Plant.genus))
)

# Create the links
links_after <- data.frame(
  source = match(after_fire_interactions$Bee.genus, nodes_after$name) - 1,
  target = match(after_fire_interactions$Plant.genus, nodes_after$name) - 1,
  value = 1
)

# Aggregate interactions for unique source-target pairs
links_after <- aggregate(value ~ source + target, data = links_after, sum)

# Create the Sankey network
library(networkD3)

color_scale <- 'd3.scaleOrdinal()
                 .domain(["Bee", "Plant"])
                 .range(["#1f78b4", "#33a02c"])'

sankey <- sankeyNetwork(
  Links = links_after,
  Nodes = nodes_after,
  Source = "source",
  Target = "target",
  Value = "value",
  NodeID = "name",
  units = "Interactions",
  fontSize = 14,
  nodeWidth = 40,
  colourScale = color_scale
)

# Render the Sankey diagram
sankey
```

Bar plot for the bee species richness.
```{r}
bee_summary_stats_df <- data.frame(
  Site.status = c("Before", "After"),
  Species.richness = c(Bee_richness_before, Bee_richness_after)
)

bee_summary_stats_df$Site.status <- factor(bee_summary_stats_df$Site.status, levels = c("Before", "After"))

ggplot(bee_summary_stats_df, aes(x = Site.status, y = Species.richness, fill = Site.status)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  geom_text(aes(label = Species.richness), vjust = -0.5, size = 4, color = "black") +
  labs(
    title = "Bee Species Richness Before and After Fire",
    x = "Site Status",
    y = "Species Richness"
  ) +
  scale_fill_manual(values = c("orange", "skyblue")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

#ggsave("bee_species_richness.png", 
       #dpi = 300,
       #units = "in",
       #bg = "white")
```
Bar plot for the plant species richness.
```{r}
plant_summary_stats_df <- data.frame(
  Site.status = c("Before", "After"),
  Species.richness = c(Plant_richness_before, Plant_richness_after)
)

plant_summary_stats_df$Site.status <- factor(plant_summary_stats_df$Site.status, levels = c("Before", "After"))

ggplot(plant_summary_stats_df, aes(x = Site.status, y = Species.richness, fill = Site.status)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  geom_text(aes(label = Species.richness), vjust = -0.5, size = 4, color = "black") +
  labs(
    title = "Plant Species Richness Before and After Fire",
    x = "Site Status",
    y = "Species Richness"
  ) +
  scale_fill_manual(values = c("orange", "skyblue")) +
  theme_minimal() +
    theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

#ggsave("plant_species_richness.png", 
       #dpi = 300,
       #units = "in",
      # bg = "white")
```

