---
title: "Project"
author: "Vincent Cortes"
date: "2024-11-19"
output: pdf_document
---
```{r load packs}
library(terra)
library(ggplot2)
library(dplyr)
library(sf)
library(maps)
library(ggspatial)
library(plotly)
library(knitr)
library(stringr)
```
Cleaning Oregon Bee Atlas data
```{r}
# Load in Oregon Bee Atlas data

oba <- read.csv("data/OBA_2018-2023.csv")
head(oba)
```
Using regular expressions to clean the species column of the OBA dataset.
```{r}
## Remove the special character
oba$Associated.plant <- str_replace_all(oba$Associated.plant, "\x92", "")

# Removing blanks
oba <- oba[oba$Associated.plant != "",]

# Removing "net"
oba$Associated.plant <- str_replace_all(oba$Associated.plant, "net", "")

# Removing yarrow
oba$Associated.plant[oba$Associated.plant == "Yarrow"] <- "Achillea millefollium"

# Removing non plant
oba <- oba[oba$Associated.plant != "Weedy yellow comp.",]

# Removing common names in ()
oba$Associated.plant <- gsub("\\s*\\(.*?\\)", "", oba$Associated.plant)

# Removing weird words
oba$Associated.plant <- str_replace_all(oba$Associated.plant, ",.*$", "")

# Removing space
oba$Associated.plant <- str_replace_all(oba$Associated.plant, "\\s+$", "")

# Removing "sp."
oba$Associated.plant <- str_replace_all(oba$Associated.plant, " sp$", " sp.")

# Removing "or"
oba$Associated.plant <- str_replace_all(oba$Associated.plant, " or .*", "")

# Removing lists
oba <- oba[!grepl(",", oba$Associated.plant),]

head(oba)
```

```{r}
all_fires <- read.csv("data/fire_data_2000-2020.csv")

head(all_fires)
all_fires_df <- all_fires[, c("FireName", "FireYear","Size_class","EstTotalAcres", "Long_DD", "Lat_DD")]
head(all_fires_df)

#filters only 2020 fires
firedata_2020 <- all_fires_df %>%
  filter(FireYear == 2020)

#remove line for missing 
firedata_2020_new <- firedata_2020[-447,]

#convert from factor to character 
firedata_2020_new$Size_class <- as.character(firedata_2020_new$Size_class)

#convert A, B, etc to 1, 2, 3, 4
firedata_2020_new$Size_class[firedata_2020_new$Size_class == "A"] <- "1"
firedata_2020_new$Size_class[firedata_2020_new$Size_class == "B"] <- "2"
firedata_2020_new$Size_class[firedata_2020_new$Size_class == "C"] <- "3"
firedata_2020_new$Size_class[firedata_2020_new$Size_class == "D"] <- "4"
firedata_2020_new$Size_class[firedata_2020_new$Size_class == "E"] <- "5"
firedata_2020_new$Size_class[firedata_2020_new$Size_class == "F"] <- "6"
firedata_2020_new$Size_class[firedata_2020_new$Size_class == "G"] <- "7"
#change them back to a factor
firedata_2020_new$Size_class <- as.factor(firedata_2020_new$Size_class)


all_fires_sf <- st_as_sf(firedata_2020_new, coords = c("Long_DD", "Lat_DD"), crs = 4326)

head(all_fires_sf)
```
Loading in the data, and building a data frame with only the columns we need. I also removed any fires that didnt have any cords, mainly just line 165.
```{r load and build df}
# Load the data as Fires
Fires <- read.csv("data/fire_data.csv")
head(Fires)

# Check column names
colnames(Fires)

# Select specific columns
fire_data_df <- Fires[, c("FireName", "FinalFireSizeAcres", "Longitude", "Latitude", "LandClass","SizeClass")]

fire_data_df
#Remove fire with no cords
fire_data_newdf <- fire_data_df[-165,]

# converts the fire data frame to a sf
fire_data_sf <- st_as_sf(fire_data_newdf, coords = c("Longitude", "Latitude"), crs = 4326)
print(fire_data_sf)

# Load Oregon county map data
oregon_map <- map_data("county", "oregon") %>% 
  select(long, lat, group, id = subregion)  # Correct column name is `region`

```

Plotting the fires based on how many acres they burned
```{r plot graph}
# Load Oregon county map data
oregon_map <- map_data("county", "oregon") %>% 
  select(long, lat, group, id = subregion)  # Correct column name is `region`

# Plot Oregon map with all fires
ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey", color = "white") +
  geom_sf(data = fire_data_sf, aes(color = SizeClass), size = fire_data_sf$SizeClass, alpha = 0.9) +
  scale_color_gradient(low = "yellow", high = "red", name = "Fire Size Class") + # Adjust the colors
  labs(title = "Distribution of Oregon Fires in 2020",
       x = "Longitude", 
       y = "Latitude") +
  theme_minimal() +
  coord_sf()


# Plot Oregon map with all fires
ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey", color = "white") +
  geom_sf(data = all_fires_sf, aes(color = Size_class), size = all_fires_sf$Size_class, alpha = 0.9) +
  labs(title = "Distribution of Oregon Fires in 2020",
       x = "Longitude", 
       y = "Latitude") +
  theme_minimal() +
  coord_sf()

Large_2020_fires_sf
```
```{r}
Large_2020_fires <- firedata_2020_new %>%
  filter(Size_class == 7)

Large_2020_fires_sf <- st_as_sf(Large_2020_fires, coords = c("Long_DD", "Lat_DD"), crs = 4326)

ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey", color = "white") +
  geom_sf(data = Large_2020_fires_sf, aes(color = Size_class), size = Large_2020_fires_sf$Size_class, alpha = 0.9) +
  labs(title = "Distribution of Oregon Fires in 2020 with level 7",
       x = "Longitude", 
       y = "Latitude") +
  theme_minimal() +
  coord_sf()

Large_2020_fires_sf
```

Loaded the bee obervations, this is from all the years, the entire data set I think. I created a dataframe with only the columns we need.
```{r add be data}
# Now lets add The bee data

#reads in the OBA bee obvercation data
Bee_Observations <- read.csv("data/observations-501373.csv")
Bee_Observations

#create a data frame
bee_observations_df <- Bee_Observations[, c("scientific_name","longitude", "latitude", "quality_grade")]

#turns data frame into sf
bee_observations_sf <- st_as_sf(bee_observations_df, coords = c("longitude", "latitude"), crs = 4326)
print(bee_observations_sf)
```

Plotted the bee observations so we can see what areas arent reached by OBA.
```{r graph bee data}
#plot bee observations on to oregon map with fires
ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey", color = "white") +
  geom_sf(data = bee_observations_sf, aes(color ="blue"), size = .1, alpha = 0.9) +
  labs(title = "Bee Species Observations in Oregon",
       x = "Longitude", 
       y = "Latitude") +
  theme_minimal() +
  coord_sf()

```
Plooted both so maybe we can pick fires that have alot ob obervations.
```{r plot fire and bee}
#plot that show Distribution of fires and bee observations
ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey", color = "white") +
  geom_sf(data = fire_data_sf, aes(color = SizeClass), size = fire_data_sf$SizeClass, alpha = 0.9) +
  scale_color_gradient(low = "yellow", high = "red", name = "Fire Size Class") + # Adjust the colors
  geom_sf(data = bee_observations_sf, color= "blue", size = .1, alpha = 0.3) +
  labs(title = "Distribution of fires and bee observations",
       x = "Longitude", 
       y = "Latitude") +
  theme_minimal() +
  coord_sf()
```

Created a 1km buffer around each fire and counted how many bee obervations occured in that buffer.
```{r}
#Now its time to go fire by fire and determin how many bee observations are present in a 1km radius of the fire origin

# create a 1 km buffer around each fire point
fire_buffers <- st_buffer(fire_data_sf, dist = 1000)  # 1000 meters = 1 km

# spatial join to find bee observations within each buffer
bee_counts <- st_join(fire_buffers, bee_observations_sf, join = st_intersects)

# Count bee observations per fire
fire_bee_counts <- bee_counts %>%
  group_by(FireName) %>%  # Group by fire name or another identifier
  summarize(BeeCount = n())  # Counts the number of bee observations

#create df
fire_bee_counts_df <- as.data.frame(fire_bee_counts)
fire_bee_counts_df

#list the top 10 fires with the most bee observations
top_fires <- bee_counts %>%
  group_by(FireName) %>%  # group by fire name 
  summarize(BeeCount = n()) %>%  # count the number of bee observations
  arrange(desc(BeeCount)) %>%  # now sort by BeeCount in descending order
  slice_head(n = 10)  # then select the top 10 fires
```
Found the fires with the top 10 bee obervations, but some of these fires are very small.
```{r}
# View the top 10 fires
print(top_fires)

top_fires_sf <- st_as_sf(top_fires, coords = c("longitude", "latitude"), crs = 4326)
print(top_fires_sf)
```

Added the bee counts to the original df
```{r merge the two df}
#left join based on the FireName column
result_df <- fire_data_newdf %>%
  left_join(fire_bee_counts_df %>% select(FireName, BeeCount), by = "FireName")

print(result_df)

```


```{r}
LionsHead_sf <- read_sf("data/LionsHead/or4472312167920200817_20200810_20210810_burn_bndy.shp")

BeachieCreek_sf <- read_sf("data/BeachieCreek/or4482112218820200816_20200810_20210810_burn_bndy.shp")

Riverside_sf <- read_sf("data/Riverside/or4504912206220200908_20190720_20210725_burn_bndy.shp")

ggplot() +
  #geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
              # fill = "lightgrey", color = "white") +
  geom_sf(data = LionsHead_sf, color = "red", alpha = 0.9) +
  geom_sf(data = BeachieCreek_sf, color = "Purple", alpha = 0.9) +
  geom_sf(data = Riverside_sf, color = "yellow", alpha = 0.9) +
  labs(
    title = "Distribution of Oregon Fires in 2020",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  coord_sf()





```