---
title: "Project"
author: "Vincent Cortes"
date: "2024-11-19"
output: pdf_document
---
```{r load packs}
library(terra)
library(ggplot2)
library(dplyr)
library(sf)
library(maps)
library(ggspatial)
library(plotly)
library(knitr)
library(stringr)
library(vegan)
library(igraph)
library(networkD3)
```
Load in the Oregon Bee Atlas data using the read.csv() function.
```{r}
# Load in Oregon Bee Atlas data

oba <- read.csv("project/data/OBA_2018-2023.csv") 
head(oba)
```
Checking how many NA Values we have in our Latitude and Longitude column. We won't be able 
to use rows with NA's so we need to drop them.
```{r}
# Initial check
sum(is.na(oba$Dec..Lat.))
sum(is.na(oba$Dec..Long.))

# Filter out the NA values
oba <- oba %>% 
  filter(!is.na(Dec..Lat.) & !is.na(Dec..Long.))

# Final Check
sum(is.na(oba$Dec..Lat.))
sum(is.na(oba$Dec..Long.))
```
Loading in the a map of Oregon for plotting our observations.
```{r, Oregon Map}
# Load Oregon county map data
oregon_map <- map_data("county", "oregon") %>% 
  select(long, lat, group, id = subregion)  # Correct column name is `region`
```

Plotted the bee observations so we can see what areas aren't reached by OBA. It also shows us the distribution of our
observations and if we have any that seem like they might be incorrect. This is just an intial visualization to explore
the observation points.
```{r graph bee data}
oba_observations_sf <- st_as_sf(oba, coords = c("Dec..Long.", "Dec..Lat."), crs = 4326)

#st_crs(oba_observations_sf)
#st_crs(oregon_map_sf) == st_crs(oba_observations_sf)

#plot bee observations on to oregon map with fires
ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey", color = "white") +
  geom_sf(data = oba_observations_sf, aes(color ="blue"), size = .1, alpha = 0.9) +
  labs(title = "Bee Species Observations in Oregon",
       x = "Longitude", 
       y = "Latitude") +
  theme_minimal() +
  coord_sf()

```

Next, we loaded in our fire data sets. These are three fires that occurred in the same area in 2020. Two of the fire
lines merged and the other fire was really close to the borders of the merged fire. These three fires are often
recorded as one large fire called the Santiam fire. We are using these because they were 3 very large fires in 2020 
that have a lot of bee observations around them. 
```{r}
# Load in each fire shape file
LionsHead_sf <- read_sf("project/data/LionsHead/or4472312167920200817_20200810_20210810_burn_bndy.shp")

BeachieCreek_sf <- read_sf("project/data/BeachieCreek/or4482112218820200816_20200810_20210810_burn_bndy.shp")

Riverside_sf <- read_sf("project/data/Riverside/or4504912206220200908_20190720_20210725_burn_bndy.shp")

# Set a common CRS 
target_crs <- st_crs(4326)

# Transform each fire to the target CRS
LionsHead_sf <- st_transform(LionsHead_sf, target_crs)
BeachieCreek_sf <- st_transform(BeachieCreek_sf, target_crs)
Riverside_sf <- st_transform(Riverside_sf, target_crs)

#check if they match
print(st_crs(LionsHead_sf) == st_crs(BeachieCreek_sf))
print(st_crs(BeachieCreek_sf) == st_crs(Riverside_sf))

# Plotting the fires on the oregon map loaded in earlier
ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey", color = "white") +
  geom_sf(data = LionsHead_sf, color = "Red", alpha = 0.9) +
  geom_sf(data = BeachieCreek_sf, color = "Red", alpha = 0.9) +
  geom_sf(data = Riverside_sf, color = "Red", alpha = 0.9) +
  labs(
    title = "Distribution of Oregon Fires in 2020",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  coord_sf()

#Now lets zoom in

# merge all fires into one
fire_combined_bbox <- st_bbox(rbind(LionsHead_sf, BeachieCreek_sf, Riverside_sf))

# Adjust the margins 
fire_combined_bbox <- fire_combined_bbox + c(-0.1, -0.1, 0.1, 0.1)  #(xmin, ymin, xmax, ymax)

# Final plotting of the zoomed in section of Oregon
ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey", color = "white") +
  geom_sf(data = LionsHead_sf, aes(color = "Lions Head"), alpha = 0.9) +
  geom_sf(data = BeachieCreek_sf, aes(color = "Beachie Creek"), alpha = 0.9) +
  geom_sf(data = Riverside_sf, aes(color = "Riverside"), alpha = 0.9) +
  scale_color_manual(
    name = "Fires",
    values = c("Lions Head" = "red", "Beachie Creek" = "purple", "Riverside" = "yellow")
  ) +
  labs(
    title = "Zoomed-In View of Oregon Fires in 2020",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  coord_sf(
    xlim = c(fire_combined_bbox["xmin"], fire_combined_bbox["xmax"]),
    ylim = c(fire_combined_bbox["ymin"], fire_combined_bbox["ymax"])
  )
```
Now we want to crop our bee observations to just the ones that are in the borders of the three fires.
We are going to paste our fires into one polygon as it is often recorded as one fire. We used the function 
st_intersection to crop our data set to the fire borders. 
```{r}
#Merge all fires into one object
all_fire_boarder <- rbind(LionsHead_sf, BeachieCreek_sf, Riverside_sf)

#Create a subset of obersvations inside the border
only_fire_bees <- st_intersection(oba_observations_sf, all_fire_boarder)
head(only_fire_bees)

only_fire_bees_df <- as.data.frame(only_fire_bees)

#merges genus and species and makes a new column
only_fire_bees_df$GenusSpecies <- paste(only_fire_bees_df$Genus, only_fire_bees_df$Species)
#create a subset
final_oba_df <- only_fire_bees_df %>% select("Year.1", "GenusSpecies","Associated.plant...genus..species", "BurnBndLat", "BurnBndLon",)

head(final_oba_df)
```

Here we plot our fires and our cropped bee observations from above on the map of Oregon
```{r, subset data}

 ggplot() +
  geom_polygon(data = oregon_map, aes(x = long, y = lat, group = group), 
               fill = "tan", color = "black") +
  geom_sf(data = LionsHead_sf, aes(color = "Lions Head"), alpha = 0.9) +
  geom_sf(data = BeachieCreek_sf, aes(color = "Beachie Creek"), alpha = 0.9) +
  geom_sf(data = Riverside_sf, aes(color = "Riverside"), alpha = 0.9) +
  geom_sf(data = only_fire_bees, aes(color = "Bee Observation"), size = 1, alpha = 0.9) +
  scale_color_manual(
    name = "Fire and Observations",
    values = c(
      "Lions Head" = "red",
      "Beachie Creek" = "purple",
      "Riverside" = "Dark Green",
      "Bee Observation" = "blue"
    ),
    breaks = c(
      "Lions Head",
      "Beachie Creek",
      "Riverside",
      "Bee Observation"
    )
  ) + 
  labs(
    title = "Oregon Santiam Fire in 2020",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.background = element_rect(fill = "white")
  ) +
  coord_sf(
    xlim = c(fire_combined_bbox["xmin"], fire_combined_bbox["xmax"]),
    ylim = c(fire_combined_bbox["ymin"], fire_combined_bbox["ymax"])
  )

#ggsave("oregon_fire_map.png", dpi = 300, width = 12, height = 8)  
```

We have a some rows with blank values or spaces that make the data difficult to use,
we will remove these to make analysis easier.
```{r}
#removed blanks
final_oba_df <- final_oba_df[final_oba_df$Associated.plant...genus..species != "",]
final_oba_df <- final_oba_df[final_oba_df$GenusSpecies != " ",]
is.na(final_oba_df)
final_oba_df
```
We also need to fix all the weird species names for the plants. There are a lot of
weird names that make analysis challenging.
```{r}
## Remove the special character
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, "\x92", "")

# Removing blanks
final_oba_df <- final_oba_df[final_oba_df$Associated.plant...genus..species != "",]

# Removing "net"
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, "net", "")

# Removing yarrow
final_oba_df$Associated.plant...genus..species[final_oba_df$Associated.plant...genus..species == "Yarrow"] <- "Achillea millefollium"

# Removing non plant
final_oba_df <- final_oba_df[final_oba_df$Associated.plant...genus..species != "Weedy yellow comp.",]

# Removing common names in ()
final_oba_df$Associated.plant...genus..species <- gsub("\\s*\\(.*?\\)", "", final_oba_df$Associated.plant...genus..species)

# Removing weird words
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, ",.*$", "")

# Removing space
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, "\\s+$", "")

# Removing "sp."
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, " sp$", " sp.")

# Removing "or"
final_oba_df$Associated.plant...genus..species <- str_replace_all(final_oba_df$Associated.plant...genus..species, " or .*", "")

# Removing lists
final_oba_df <- final_oba_df[!grepl(",", final_oba_df$Associated.plant...genus..species),]

head(final_oba_df)
```

Here we are creating two columns to sort the plant and bee species by genus and species.
When we make our networks it will make it easier to have just the genus and species sorted. We will start
with identifying the species resolution then we will drop the values that are missing a genus or species.
```{r}
# Initialize the empty columns to NA values
final_oba_df$Plant.resolution <- NA
final_oba_df$Bee.resolution <- NA

# Loop through the data frame and then check if there is only one word in the
# plant column. If there's one word it will be the Genus, otherwise it's species.
for (i in 1:nrow(final_oba_df)) {
  if (str_count(final_oba_df$Associated.plant...genus..species[i], "\\S+") == 1) {
    final_oba_df$Plant.resolution[i] <- "Genus"
  }
  else {
    final_oba_df$Plant.resolution[i] <- "Species"
  }
}

# Similar to above we will loop through our data frame and determine if the Bee resolution is
# Genus or Species
for (i in 1:nrow(final_oba_df)) {
  if (str_count(final_oba_df$GenusSpecies[i], "\\S+") == 1) {
    final_oba_df$Bee.resolution[i] <- "Genus"
  }
  else {
    final_oba_df$Bee.resolution[i] <- "Species"
  }
}

# Check our resolutions
table(final_oba_df$Bee.resolution)
table(final_oba_df$Plant.resolution)

# Now we want to filter our plant species to the rows that contain 
# just the genus
final_oba_df$Plant.genus <- ifelse(
  final_oba_df$Plant.resolution %in% c("Genus", "Species"),
  word(final_oba_df$Associated.plant...genus..species, 1), NA)

# Check the plant genus
table(final_oba_df$Plant.genus)

# Now we will filter our bee species to both Genus and species
final_oba_df$Bee.genus <- ifelse(
  final_oba_df$Bee.resolution %in% c("Genus", "Species") & str_count(final_oba_df$GenusSpecies, "\\S+") > 1, word(final_oba_df$GenusSpecies, 1, 2), NA)

# Check our bee genus 
table(final_oba_df$Bee.genus)
```

Drop the NA values to get our final clean data set with the correct bee and plant resolutions.
```{r}
final_oba_df <- final_oba_df %>% 
  filter(!is.na(Plant.genus) & !is.na(Bee.genus))

final_oba_df
```

Add the site status column to our final clean data frame. We want the site status to be "Before" if
the year of the observation was 2020 and before and "After" if the observation was after 2020.
```{r}
# Initialize an empty site status column
final_oba_df$Site.status <- NA

# Loop through the oba data frame
for (i in 1:nrow(final_oba_df)) {
  if (final_oba_df$Year.1[i] <= 2020) {
    final_oba_df$Site.status[i] <- "Before"
  }
  else {
    final_oba_df$Site.status[i] <- "After"
  }
}

# Check our new data frame
head(final_oba_df)
```

Creating a species interaction network for before the fires.
```{r}
# Filter data for 'before' fire interactions
before_fire_interactions <- final_oba_df[final_oba_df$Site.status == "Before", ]

# Create the nodes
nodes_before <- data.frame(
  name = unique(c(before_fire_interactions$Bee.genus, before_fire_interactions$Plant.genus))
)

# Create the links
links_before <- data.frame(
  source = match(before_fire_interactions$Bee.genus, nodes_before$name) - 1,
  target = match(before_fire_interactions$Plant.genus, nodes_before$name) - 1,
  value = 1
)

# Aggregate interactions for unique source-target pairs
links_before <- aggregate(value ~ source + target, data = links_before, sum)

# Create the Sankey network
library(networkD3)

# Define a color scale for nodes
color_scale <- 'd3.scaleOrdinal()
                 .domain(["Bee", "Plant"])
                 .range(["#1f78b4", "#33a02c"])'

sankey_before <- sankeyNetwork(
  Links = links_before,
  Nodes = nodes_before,
  Source = "source",
  Target = "target",
  Value = "value",
  NodeID = "name",
  units = "Interactions",
  fontSize = 14,
  nodeWidth = 40,
  colourScale = color_scale
)

# Here is the network
sankey_before
```

Now we want to calculate some statistics to simualte. We want to calculate the species richness before
and after the fires and then take the difference in richness for both plants and bees.
```{r}
# Richness for bees
Bee_richness_before <- length(unique(final_oba_df$Bee.genus[final_oba_df$Site.status == "Before"]))
Bee_richness_after <- length(unique(final_oba_df$Bee.genus[final_oba_df$Site.status == "After"]))
Bee_richness_change <- Bee_richness_after - Bee_richness_before

# Richness for plants
Plant_richness_before <- length(unique(final_oba_df$Plant.genus[final_oba_df$Site.status == "Before"]))
Plant_richness_after <- length(unique(final_oba_df$Plant.genus[final_oba_df$Site.status == "After"]))
Plant_richness_change <- Plant_richness_after - Plant_richness_before
```

Now we will conduct our simulation using bootstrapping. Our sample size for our "after" group is really small 
so we will use resampling and then a density histogram to simulate what our hypothesis could look
like if there was more data available.
```{r}
# Observed difference in richness
bee_observed_diff <- Bee_richness_after - Bee_richness_before

# Resampling procedure because our "after" group is extremmely small
set.seed(42)  # Set a seed for reproducibility
num_iterations <- 10000  # Increase iterations for a larger distribution

# Perform simulation with shuffled Site.status
bee_null_distribution <- replicate(num_iterations, {
  # Shuffle Site.status
  shuffled_status <- sample(final_oba_df$Site.status)
  
  # Assign the shuffled status to a new column
  final_oba_df$Shuffled.status <- shuffled_status
  
  # Sample from the shuffled "Before" and "After" groups
  bee_sampled_before <- final_oba_df$Bee.genus[final_oba_df$Shuffled.status == "Before"]
  bee_sampled_after <- final_oba_df$Bee.genus[final_oba_df$Shuffled.status == "After"]
  
  # Calculate richness
  bee_richness_before <- length(unique(bee_sampled_before))
  bee_richness_after <- length(unique(bee_sampled_after))
  
  # Return the difference in richness
  bee_richness_after - bee_richness_before
})

# Compare observed difference with null distribution
p_value <- mean(abs(bee_null_distribution) >= abs(observed_diff))
print(paste("p-value: ", p_value))

# Plot the null distribution
bee_null_df <- data.frame(bee_null_distribution)

ggplot(bee_null_df, aes(x = bee_null_distribution)) +
  geom_density(fill = "lightblue", alpha = 0.7) +
  geom_vline(aes(xintercept = bee_observed_diff), color = "red", linetype = "dashed", size = 1) +
  labs(title = "Null Distribution of Bee Species Richness",
       x = "Difference in Richness",
       y = "Density") +
  theme_minimal() +
   theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

#ggsave("simulated_bee_species_richness.png", 
       #plot = bee_plot,
       #dpi = 300,
       #units = "in",
      #bg = "white")
```

Here we conduct our simulation for the plants.
```{r}
# Observed difference in richness
plant_observed_diff <- Plant_richness_after - Plant_richness_before

# Resampling procedure because our "after" group is extremmely small
set.seed(42)  # Set a seed for reproducibility
num_iterations <- 10000  # Increase iterations for a larger distribution

# Perform simulation with shuffled Site.status
plant_null_distribution <- replicate(num_iterations, {
  # Shuffle Site.status
  shuffled_status <- sample(final_oba_df$Site.status)
  
  # Assign the shuffled status to a new column
  final_oba_df$Shuffled.status <- shuffled_status
  
  # Sample from the shuffled "Before" and "After" groups
  plant_sampled_before <- final_oba_df$Plant.genus[final_oba_df$Shuffled.status == "Before"]
  plant_sampled_after <- final_oba_df$Plant.genus[final_oba_df$Shuffled.status == "After"]
  
  # Calculate richness
  plant_richness_before <- length(unique(plant_sampled_before))
  plant_richness_after <- length(unique(plant_sampled_after))
  
  # Return the difference in richness
  plant_richness_after - plant_richness_before
})

# Compare observed difference with null distribution
p_value <- mean(abs(plant_null_distribution) >= abs(observed_diff))
print(paste("p-value: ", p_value))

# Plot the null distribution
plant_null_df <- data.frame(plant_null_distribution)
ggplot(plant_null_df, aes(x = plant_null_distribution)) +
  geom_density(fill = "lightblue", alpha = 0.7) +
  geom_vline(aes(xintercept = plant_observed_diff), color = "red", linetype = "dashed", size = 1) +
  labs(title = "Null Distribution of Plant Species Richness",
       x = "Difference in Richness",
       y = "Density") +
  theme_minimal() +
   theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

#ggsave("simulated_plant_species_richness.png", 
       #dpi = 300,
       #units = "in",
      #bg = "white")
```

Species interaction network for after the fires.
```{r}
# Filter data for 'before' fire interactions
after_fire_interactions <- final_oba_df[final_oba_df$Site.status == "After", ]

# Create the nodes
nodes_after <- data.frame(
  name = unique(c(after_fire_interactions$Bee.genus, after_fire_interactions$Plant.genus))
)

# Create the links
links_after <- data.frame(
  source = match(after_fire_interactions$Bee.genus, nodes_after$name) - 1,
  target = match(after_fire_interactions$Plant.genus, nodes_after$name) - 1,
  value = 1
)

# Aggregate interactions for unique source-target pairs
links_after <- aggregate(value ~ source + target, data = links_after, sum)

# Create the Sankey network
library(networkD3)

color_scale <- 'd3.scaleOrdinal()
                 .domain(["Bee", "Plant"])
                 .range(["#1f78b4", "#33a02c"])'

sankey <- sankeyNetwork(
  Links = links_after,
  Nodes = nodes_after,
  Source = "source",
  Target = "target",
  Value = "value",
  NodeID = "name",
  units = "Interactions",
  fontSize = 14,
  nodeWidth = 40,
  colourScale = color_scale
)

# Render the Sankey diagram
sankey
```

Bar plot for the bee species richness.
```{r}
bee_summary_stats_df <- data.frame(
  Site.status = c("Before", "After"),
  Species.richness = c(Bee_richness_before, Bee_richness_after)
)

bee_summary_stats_df$Site.status <- factor(bee_summary_stats_df$Site.status, levels = c("Before", "After"))

ggplot(bee_summary_stats_df, aes(x = Site.status, y = Species.richness, fill = Site.status)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  geom_text(aes(label = Species.richness), vjust = -0.5, size = 4, color = "black") +
  labs(
    title = "Bee Species Richness Before and After Fire",
    x = "Site Status",
    y = "Species Richness"
  ) +
  scale_fill_manual(values = c("orange", "skyblue")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

#ggsave("bee_species_richness.png", 
       #dpi = 300,
       #units = "in",
       #bg = "white")
```
Bar plot for the plant species richness.
```{r}
plant_summary_stats_df <- data.frame(
  Site.status = c("Before", "After"),
  Species.richness = c(Plant_richness_before, Plant_richness_after)
)

plant_summary_stats_df$Site.status <- factor(plant_summary_stats_df$Site.status, levels = c("Before", "After"))

ggplot(plant_summary_stats_df, aes(x = Site.status, y = Species.richness, fill = Site.status)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  geom_text(aes(label = Species.richness), vjust = -0.5, size = 4, color = "black") +
  labs(
    title = "Plant Species Richness Before and After Fire",
    x = "Site Status",
    y = "Species Richness"
  ) +
  scale_fill_manual(values = c("orange", "skyblue")) +
  theme_minimal() +
    theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

#ggsave("plant_species_richness.png", 
       #dpi = 300,
       #units = "in",
      # bg = "white")
```

